---
output: html_document
author: Vijayvithal
title: Bangalore demographics
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```
# Introduction

This document uses the voterlist data for conducting a demographic analysis of Bangalore. While the data is extensive, ECI is known to have poor data entry and maintainance habits.

```{r}
ages<-read.delim("ages_by_ward.txt")
plot(unique(ages$Age),tapply(ages$Voters,ages$Age,mean,omit=TRUE),ylab="Voter count", xlab="Age",main="Voters In Bangalore")
```
We see that the age distribution peaks at around 30-35 years and then gradually falls until 80.
We will see if we can identify the pensioners paradise, student enclaves, working singles, working families and migrant areas of bangalore based on the voter data.
The different groupings are defined as follows.

* Students: Age group 18 to 22 yrs
* Working singles: 22 to 26(Women) 28(men) we will take an average of 27 as the upper bound
* Working families: 27 to 58
* Retired Citizens: 58 to 62
* Senior Citizens 62+
* Migrant Areas:Large working population with a spike around retirement age when parents relocate to be there with the children.

Since the absolute ward population can differ we will represent each of this group as a percentage of the total ward population.
```{r}
#df<- data.frame(ward=character(),total=numeric(),students=numeric(),singles=numeric(),families=numeric(),retired=numeric(),seniors=numeric())
#  for (ward in unique(ages$Ward)){
#  av<-ages[ages$Ward==ward,]
#  w.sum<-sum(av$Voters)
#  students<-av[av$Age<23,]
#  singles<-av[av$Age>=23 & av$Age<27,]
#  familiesit<-av[av$Age>=27 & av$Age<40,]
#   familiesmix<-av[av$Age>=40 & av$Age<50,]
#   familiesnit<-av[av$Age>=50 & av$Age<58,]
#   retired<-av[av$Age>=58 & av$Age<62,]
#   seniors<-av[av$Age>=62,]
#   w.student<-100*sum(students$Voters)/w.sum
#   w.single<-100*sum(singles$Voters)/w.sum
#   w.familyIT<-100*sum(familiesit$Voters)/w.sum
#   w.familyMIX<-100*sum(familiesmix$Voters)/w.sum
#   w.familyNonIT<-100*sum(familiesnit$Voters)/w.sum
#   
#   w.retired<-100*sum(retired$Voters)/w.sum
#   w.senior<-100*sum(seniors$Voters)/w.sum
#   df<-rbind(df,data.frame(ward=ward,total=w.sum,students=w.student,singles=w.single,ITfamilies=w.familyIT,SITfamilies=w.familyMIX,NonITfamily=w.familyNonIT,retired=w.retired,seniors=w.senior))
#  }

####
# The code above was used to generate a csv file to which data related to ward/ac/zone were added. We will directly read in the csv file
####
ag_data<- read.csv("age_df.csv")
quant_it<-quantile(ag_data$ITfamilies,probs=(0:10)/10)
```


Which groups contain interesting data?

The standard deviation for students and retired groups is very low, we find high deviation for New families and senior citizens group.
Analysing these two sets of data...

``` {r}


tmp <- data.frame(name=character(),sd=numeric())
              
for (var in colnames(ag_data)) {
if(is.numeric(ag_data[[var]])) {
x <- quantile(ag_data[[var]],probs=(0:10)/10)
sdd <- sd(ag_data[[var]])
  
tmp <- rbind(tmp,data.frame(name=var,sd=sdd))
             
}
  
}

tmp

#,'10'=numeric(),'20'=numeric(),'30'=numeric(),'40'=numeric(),'50'=numeric(),'60'=numeric(),'70'=numeric(),'80'=numeric(),'90'=numeric(),'100'=numeric())

#,'10'=x['10%'],'20'=x['20%'],'30'=x['30%'],'40'=x['40%'],'50'=x['50%'],'60'=x['60%'],'70'=x['70%'],'80'=x['80%'],'90'=x['90%'],'100'=x['100%']))


```


Potential issues with the analysis.

* Typically college students staying in hostels do not register for a voter card unless there is a dedicated voter registration drive in that college.
* Migrant population face discrimination from the government machinery and may not show up in full strength on the voter list.

# Initial tests

Where are the Young families settling down?
The system tells us that young families are settling down in 

* Mahadevapura(5 wards)
* Bommanahalli(4 wards)
* Dasarhalli (4 wards)

Along with a few wards in the neighbourhood of the above wards.
Checking on the map we see 

* Mahadevapura contains various IT parks on Outer ring road between Kr Puram and Sarjapur, ITPL, Bagmane tech park.
* Dasarhalli contains the Peenya Industrial Area
* Bommanahalli is close to Electronic city (confirm?)

```{r}
quant_it<-quantile(ag_data$ITfamilies,probs=(0:10)/10)
ag_data[ag_data$ITfamilies>quant_it['90%'],2:5]
```

Which are the "Pensioners paradise?"
The data reports old bangalore areas as the areas where citizens above 60 years are clustered.
```{r}
auant_it<-quantile(ag_data$seniors,probs=(0:10)/10)
ag_data[ag_data$seniors>auant_it['90%'],2:5]
```

Measuring the quartiles across each ward we get
```{r}
age_quartile<-function(ages) {
  retval<- c()
  for (age in unique(ages$Age)){
  av<-ages[ages$Age==age,]
  q<-quantile(av$Voters)
retval<-rbind(retval,q)
  #print q
  }
   retval
}
r<-age_quartile(ages)
df_r=as.data.frame(r)
plot(y=df_r$`100%`,x=18:(18+112),type='l',yaxp=c(0,3500,10))
lines(y=df_r$`25%`,x=18:(18+112),col="red")
lines(y=df_r$`50%`,x=18:(18+112),col="green")
lines(y=df_r$`75%`,x=18:(18+112),col="blue")
lines(y=df_r$`00%`,x=18:(18+112),col="cyan")


```


