---
title: "Voter list analysis"
author: "Vijayvithal"
date: "8 September 2015"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

This is an analysis of the voters list for Entire Bangalore city.
The database consists of 4 years worth of voter records, each year has over 6 million records. For each voter the following data is captured.

* AC[0-9]+ This is a combination of the AC and the part number
* [0-9]+ This is a 3 digit code
* [0-9]+ This is the serial number
* [A-Z]+[0-9]+ This is the voter ID Number
* [A-Z]+ This is the voters Name
* [FM] Voters Gender
* [0-9]+ Voters Age
* [A-Z]+ Voters Relative
* [FH] Relationship with Relative

# Preprocessing.
R is not able to read the data and crashes with a memory limit message. So as a first step we will split each year's data into separate AC wise data files.
```{r}
#Checking which version of Perl is installed
Sys.which('perl')
```

```{r test-perl, engine='perl'}
use strict;
use warnings;
use Archive::Zip;
my $zip = Archive::Zip->new;
return if(-d "csvfiles"); # This is a lengthy process, skip it if the script has been run before.
mkdir "csvfiles";
my @files=glob("voter_record/*.zip");
my $prevfile="ViJaY";
foreach my $file(@files){
  my $outfile;
  $file=~/.*\/([^\/]*).zip/;
  my $dirname="csvfiles/".$1;
  mkdir $dirname;
  $zip->read ($file);
  foreach my $member ($zip->members){
    next if $member->isDirectory;
  $member->extractToFileNamed("Archive.csv");
  open(FILE,"<Archive.csv");
  while (<FILE>){
    s/,/./g;
    s/^\s*//;
    next if (/^\s*$/);
    my @fields=split("\t");
    $fields[0]=~/^([A-Z]*...)/i;
    my $ACname=$1;
    if($prevfile ne $ACname) {
      open (OFILE,">$dirname/$ACname.csv") or die "Unable to open file $dirname/$ACname.csv:$1 with $fields[0] fields @fields";
      $prevfile=$ACname;
    }
    print OFILE $_;
    
  }
  close(FILE);
  }
}
# By this point we hav a folder called csvfiles which has a folder for each dataset further divided by AC's
```

